Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence.

### 2022-08-12 reports/manuscript_v4.Rmd
mac version

---
title: "Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: elsevier_vancouver.csl
---

@@

NEEDS UPDATING

@@

There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@Bernstein]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. The impact of endemic zoonoses meanwhile remains underestimated [@Maudlin]. Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@Halliday; @Molyneux; @judson_zoonoses_2021]. West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016]. Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015]. Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

@@

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@Bowler; @Boakes]. These sampling biases can gravely distort produced species distribution models [@Beck]. Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@Gibb_clover]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

@@

Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented. GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@GeoNames] based on place-names and maps presented in the study. All locations were converted to decimal degrees for internal consistency. The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity. The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1). For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

@@

To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@Database of Global Administrative Areas]. We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above. The sf package in the R statistical language was used to handle shapefiles and geographic data [@Pebesma]. Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area. For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@SEDAC].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@COPERNICUS]. The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.

@@

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{2}
\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{3}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{4}
\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{5}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}
$$

@@

Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations. To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@Gibb_data_proliferation]. CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi. An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@Gibb_clover_data].

@@

Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015). Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights. Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g. IUCN, GBIF). The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@Boakes]. We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset. Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@Harvey].

We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@Beck]. Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M. musculus* and *R. rattus*) and the hazard of endemic zoonosis outbreaks (e.g. Lassa mammarenavirus). Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges. For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@Basinski]. Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.

Rodent trapping studies provide data on both presence and absence which are not available from curated datasets. Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@Vaclavik]. Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.

Currently available consolidated datasets on host-pathogen associations (e.g. CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@Gibb_data_proliferation]. The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range. Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range. Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R. rattus* and Q fever (*Coxiella burnetii*). Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@Wilkinson].

Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics. This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet_calvet_2016]. Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.

Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies. Land use change and population density have changed dramatically over this period in West Africa [@hermann]. We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005. Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to significantly change.
We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data. Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence. However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@Foster].

Future rodent trapping studies should be targeted towards regions that are currently under-studied. Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence. Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis. Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.

>>>>

---
title: "Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
  # pdf_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: harvard.csl
latex_engine: xelatex
---

@@

Rodents are important globally distributed reservoirs of known and novel zoonotic pathogens. Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of endemic zoonoses spillover events. These changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk. However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are biased by the inclusion of limited data sources and spatial biases in sampled locations. Here, we synthesise data from West Africa from 127 rodent trapping studies, conducted between 1964-2021, including over 72,000 individual rodent detections to more comprehensively characterise the range and presence of important zoonotic pathogen host species in this region. We identify that rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and calculate the discrepancies between sampling in GBIF and included studies. For five regionally important zoonotic pathogens (Arenaviridae spp., Borrelia spp., *Lassa mammarenavirus*, Leptospira spp. and *Toxoplasma gondii*) we identify host-pathogen associations that are not reported in the consolidated CLOVER dataset. These host-pathogen asssociation omissions may bias studies investigating the current risk of endemic zoonoses and the effect of land use and climate change on future novel pathogen emergence. Finally, we find that for these five pathogen groups, the proportion of a rodent hosts range that has been sampled remains small. A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk. In the interim, future studies must incorporate a measure of the current sampling biases. The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support a more complete understanding of the hazard of zoonotic spillover events.

@@

There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_nodate]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. The impact of endemic zoonoses meanwhile remains underestimated [@maudlin_neglected_2009]. Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@halliday_endemic_2015; @molyneux_zoonoses_2011; @judson_zoonoses_2021]. West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016]. Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015]. Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

@@

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@bowler_temporal_nodate; @boakes_distorted_2010]. These sampling biases can gravely distort produced species distribution models [@beck_spatial_2014]. Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@gibb_viralemergenceclover_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

@@

Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented. GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study. All locations were converted to decimal degrees for internal consistency. The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity. The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1). For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

@@

To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@database_of_global_administrative_areas_gadm_2022 of Global Administrative Areas]. We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above. The sf package in the R statistical language was used to handle shapefiles and geographic data [@pebesma_simple_2018]. Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area. For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@socioeconomic_data_and_applications_center_gridded_2021].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.

@@

$$

\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}

\tag{2}
\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}

\tag{3}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}

\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}

\tag{5}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}

$$

@@

Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations. To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021]. CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi. An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@gibb_viralemergenceclover_2021].

@@

Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (Han et al. 2015). Here we have synthesised data from 126 rodent trapping studies containing information on more than 76,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries. Locations studied are complementary to curated datasets (e.g. IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets (Boakes et al., 2010). We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset. However, pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots (Harvey and Holmes, 2022).

We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases (Beck et al., 2014). Relative trapping effort was biased towards increased effort in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion of non-native rodent species (M. musculus and R. rattus) and the risk of endemic zoonosis outbreaks (e.g. Lassa mammarenavirus). Much of West Africa remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges. For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana (Basinski et al., 2021). Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.

Rodent trapping studies provide geographic and temporally contextualised data on both species detection and non-detection which are not available from curated datasets. Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences (Václavík and Meentemeyer, 2009). Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk (Plowright et al., 2019).

Currently available consolidated datasets on host-pathogen associations (e.g. CLOVER, EID2 and GMPD2) do not include spatial or temporal components (Gibb, Gregory F. Albery, et al., 2021a). The current synthesis of rodent trapping studies has highlighted that pathogens have been sparsely sampled within a hosts range. Current zoonosis risk models dependent on these sources of data are therefore not able to incorporate spatial heterogeneity in pathogen prevalence across the host range. Additional uncertainty in current models of zoonotic disease risk arises from observed host-pathogen associations that have not been reported in these consolidated datasets. For example, Hylomyscus pamfi infected with Lassa mammarenavirus and R. rattus infected with Coxiella burnetiid, will not be included when solely dependent on consolidated host-pathogen datasets. Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses (Wilkinson et al., 2019).

Few studies stratified detection and non-detection of hosts or pathogen prevalence by time, therefore limiting inference of changes in host and pathogen dynamics. This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space (Fichet-Calvet et al., 2016). Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.

Finally, due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies. Land use change and population density have changed dramatically over this period in West Africa (Herrmann et al., 2020). We attempted to mitigate against this by using the median year of trapping as a reference point to understand the spatial and land use biases in trapping activity. It is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the land use and human population variables in 2005. Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change (Beck et al., 2014).

We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data. Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence. Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication (Foster and Deardorff, 2017).

Future rodent trapping studies should be targeted towards regions that are currently under-studied. Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence. Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis. Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events. Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement curated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
